---
title: Curation of biobank ssi from Stabl
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

## Data basic information

This curates the biobank ssi data from [Stabl](https://github.com/gregbellan/Stabl/tree/main)


Surgical Site Infections (SSI)

Training

- Outcome: Control (`77`) Vs. SSI (`16`)
- CyTOF: `93` samples — `1125` biomarkers
- Proteomics: `91` samples — `721` biomarkers

```{r setup, message=FALSE}
# Load libraries
library(here)
library(tidyverse)
library(MultiAssayExperiment)
library(mixOmics)
# Custom functions to use
plot_pca <- function(df, title=NULL, ...) {
  if (is.null(title)) stop("Need to supply title")
  pcs <- pca(df) 
  plotIndiv(pcs, title=title, ...)
}

plot_hist <- function(df, ...) {
  x <- df %>% 
    select_if(is.numeric) %>%
    as.matrix()
  
  # Plot it
  hist(x=x, ...)
}

plot_response_dist <- function(df, y_col, width=0.5, palette="Dark2", ...) {
  df |> 
  mutate({{y_col}} := factor({{y_col}})) %>%
  ggplot(aes(x = {{y_col}}, fill={{y_col}}), ...) +
  geom_bar(width = width) +
  labs(...) +
  scale_fill_brewer(palette = palette) +
  theme_bw()
}



# Parameters to use
data_dir <- here("curation_data/stabl/stabl_data/biobank_ssi")
cytof_path <- here(data_dir, "CyTOF.csv")
proteomics_path <- here(data_dir, "Proteomics.csv")
outcome_path <- here(data_dir, "outcome.csv")
```


## Sanity Checks

Then, we could first load the data and inspect each of them

```{r read-data, message=F}
# Loading the data
cytof <- read_csv(cytof_path)
proteomics <- read_csv(proteomics_path)
outcome <- read_csv(outcome_path)
```

Cytof looks like this

```{r}
# First 6 rows are like
head(cytof)
```
Then, we could perform pca on the data to see if any outlier, and plot histogram to check overall distribution of all the numeric variables

```{r}
# Check outlier
plot_pca(cytof, title="CyTOF")
plot_hist(cytof, main="Histogram of CyTOF distribution")
```


Outcome looks like this

```{r}
# First 6 rows
head(outcome)
head(outcome) |> summary()
```

Then, count distribution of classes in the outcome

```{r}
# Plot the distribution of classes in the outcome
plot_response_dist(outcome, model1b, x="Model1b", fill = "Model1b",
                   title="Distribution of classes in the outcome")
```


summary of proteomics

```{r}
# Show first 6 rows
head(proteomics)
```

Proteomics has one identifier column `r colnames(proteomics)[1]`, where rest are numeric. So, we could see overall distribution of these biomarker variables.

```{r}
# PCA on it to see if any noticeable outlier
plot_pca(proteomics, title = "Proteomics", )
# Histogram to see overal distributions of values of each biomarker
plot_hist(proteomics, main="Histogram of Proteomics biomarkers")
```

Notice, there is a common identifier among the 3 objects, so we could use it to wrap out data into MultiAssayExperiment format.

```{r}
intersect(colnames(cytof), intersect(colnames(outcome), colnames(proteomics)))
```

## MultiAssayExperiment conversion

Before conversion, we need to check dimension of all the data and see if they could match

```{r}
all_data <- list(cytof=cytof, proteomics=proteomics, outcome=outcome)
# Check dimension
lapply(all_data, dim)
# Now noticed we only had 91 in proteomics, so we need to reduce the size of 
# our data from 91 to 93
common_samples <- Reduce(intersect, sapply(all_data, function(x) x$sampleID)) 
# Then we could subset those data only
common_data <- lapply(all_data, function(x) x %>% filter(sampleID %in% common_samples))
# Then extract the X and y component
# For MAE we required dimensions to Ni x M, where M is number of obs
X <- common_data[1:2]  |> lapply(function(omic){
  # For each omic, we need to put the sampleID to be rowname
  # and transpose it for the MAE req
  mat <- omic |>
        tibble::column_to_rownames(var="sampleID") |>
        t()
  return(mat)
  
})
Y <- common_data[[3]]
```

```{r mae-conversion}
# Wrap it to MAE constructor, notice its named "mae_"
mae <- MultiAssayExperiment(experiments = X, metadata = Y)
# And add the response to a slot of the mae
mae$response <- Y$model1b
# Where it looks like this
mae
```

Saving this to disk

> TODO: fix here

```{r}
# TODO: save the MultiAssayExperiment to disk

```











