---
title: "GSE71669"
author: "Amrit Singh"
date: "2023-07-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(GEOquery)
library(ENmix)
library(HDF5Array)
library(MultiAssayExperiment)
library(annotate)
library(hugene10sttranscriptcluster.db)
library(tidyverse)
library(mixOmics)
library(immunedeconv)
```

## bladder cancer dataset
 - gene expression + methylation
 - [GSE71669](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE71669)

```{r}
gds <- getGEO("GSE71669")
```

# Gene expression
- [HuGene-1_0-st] Affymetrix Human Gene 1.0 ST Array [transcript (gene) version]

## import mrna exprs

```{r}
mrna_exprs <- exprs(gds$`GSE71669-GPL6244_series_matrix.txt.gz`)
```

## QC on mrna datasets

```{r}
pcs <- pca(t(mrna_exprs))
plotIndiv(pcs, title = "mrna")
```
> no clear outliers

## import sample metadata

```{r}
mrna_pdata <- pData(gds$`GSE71669-GPL6244_series_matrix.txt.gz`)
```

## Create response variable (IMPORTANT!)

```{r}
stage <- mrna_pdata$`tstage:ch1`
stage[grep("Ta", stage)] <- "non-invasive bladder cancer"
stage[!(stage %in% "non-invasive bladder cancer")] <- "invasive bladder cancer"

## add response to pdata
mrna_pdata$response <- factor(stage, levels = c("non-invasive bladder cancer", "invasive bladder cancer"))

## do any samples repeat
any(table(mrna_pdata$title) > 1) # nope
```


## check if sample ordering matches between mrna_exprs and mrna_pdata

```{r}
all(rownames(mrna_pdata) == colnames(mrna_exprs))

## rename sample to patient id
rownames(mrna_pdata) <- colnames(mrna_exprs) <- mrna_pdata$title
```


## import mrna annotation

```{r}
mrna_ann <- data.frame(probeids = rownames(mrna_exprs),
                       gensym = getSYMBOL(rownames(mrna_exprs), "hugene10sttranscriptcluster"))
mrna_ann <- na.omit(mrna_ann)
```

## only keep probes with non-NA Gene Symbols

```{r}
mrna_exprs <- mrna_exprs[mrna_ann$probeids, ]
hist(mrna_exprs)
```

> removed probe with genesym=NA, data is already log2 transformed

## Methylation
- 	Illumina HumanMethylation27 BeadChip (HumanMethylation27_270596_v.1.2)

### import cpg data

```{r}
cpg_exprs <- exprs(gds$`GSE71669-GPL8490_series_matrix.txt.gz`)
```

### QC on cpg data

```{r}
par(mfrow = c(1, 2))
hist(cpg_exprs)
plot(rowMeans(cpg_exprs))
```

#### remove negative values

```{r}
## remove cpgs with negative values
cpg_exprs[cpg_exprs < 0] <- NA
cpg_exprs <- na.omit(cpg_exprs)
```

> data seems to be beta values [0,1]


#### outlier detection

```{r}
hist(cpg_exprs)

pcs <- pca(t(cpg_exprs))
plotIndiv(pcs, title="cpgs")
```

## import sample metadata

```{r}
cpg_pdata <- pData(gds$`GSE71669-GPL8490_series_matrix.txt.gz`)
```

## check if sample ordering matches between mrna_exprs and mrna_pdata

```{r}
all(rownames(cpg_pdata) == colnames(cpg_exprs))

## rename sample to patient id
rownames(cpg_pdata) <- colnames(cpg_exprs) <- cpg_pdata$title
```

## import cpg annotation

```{r}
cpg_ann <- gds$`GSE71669-GPL8490_series_matrix.txt.gz`@featureData@data
cpg_ann <- cpg_ann[rownames(cpg_exprs), ]
```

## Estimate cell-type frequencies

### mRNA

```{r}
mrna_exprs_avg0 <- mrna_exprs %>% 
  as.data.frame() %>% 
  mutate(probeids = rownames(.)) %>% 
  inner_join(mrna_ann, "probeids") %>% 
  dplyr::select(-probeids) %>% 
  group_by(gensym) %>% 
  summarise(across(everything(), mean))
mrna_exprs_avg <- as.matrix(mrna_exprs_avg0[, -1])
rownames(mrna_exprs_avg) <- mrna_exprs_avg0$gensym

xCell.data <- immunedeconv::xCell.data
cc_mrna <- immunedeconv::deconvolute(mrna_exprs_avg, "xcell")
cc_mrna$cell_type <- gsub(" ", ".", cc_mrna$cell_type)

pcs <- pca(t(cc_mrna[,-1]))
plotIndiv(pcs)
```

> outlier detected ("GSM1838846")

#### plot immune cells after removing outlier

```{r}
cc_mrna %>% 
  gather(sample, fraction, -cell_type) %>%
  # plot as stacked bar chart
  ggplot(aes(x = sample, y = fraction, fill = cell_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_x_discrete(limits = rev(levels(cc_mrna)))

```

> outlier detected ("GSM1838846")

### CpGs 

```{r}
cc_cpg=estimateCellProp(userdata=cpg_exprs,
                          refdata="FlowSorted.Blood.450k",
                          nonnegative = TRUE)

pcs <- pca(cc_cpg[,-1])
plotIndiv(pcs)
```

> combine cell-type dataset and compare

## merge metadata

```{r}
colnames(mrna_pdata) <- paste("mrna", colnames(mrna_pdata), sep="_")
colnames(mrna_pdata)[colnames(mrna_pdata) == "mrna_title"] <- "title"
colnames(mrna_pdata)[colnames(mrna_pdata) == "mrna_response"] <- "response"
colnames(cpg_pdata) <- paste("cpg", colnames(cpg_pdata), sep="_")
colnames(cpg_pdata)[colnames(cpg_pdata) == "cpg_title"] <- "title"

## union
metadata_all <- mrna_pdata %>% 
  full_join(cpg_pdata, "title") %>% 
  as.data.frame()
rownames(metadata_all) <- metadata_all$title

## intersection
metadata_complete <- mrna_pdata %>% 
  inner_join(cpg_pdata, "title") %>% 
  as.data.frame()
rownames(metadata_complete) <- metadata_complete$title

write.csv(metadata_all, "metadata_all.csv")
write.csv(metadata_complete, "metadata_complete.csv")
```

## merge cells datasets

```{r}
## reformat cc_mrna data
cc_mrna2 <- as.matrix(cc_mrna[,-1])
rownames(cc_mrna2) <- cc_mrna$cell_type
cc_mrna2 <- t(cc_mrna2) %>% 
  as.data.frame() %>% 
  mutate(Sample_Name = rownames(.))

intersect(colnames(cc_mrna2), colnames(cc_cpg)) ## no cell-type names match
cc_exprs_all = data.frame(full_join(cc_mrna2, cc_cpg, "Sample_Name"))
rownames(cc_exprs_all) <- cc_exprs_all$Sample_Name


## cc
cc_exprs_all <- cc_exprs_all[metadata_all$title, setdiff(colnames(cc_exprs_all), "Sample_Name")]
cc_exprs_complete <- cc_exprs_all[metadata_complete$title, setdiff(colnames(cc_exprs_all), "Sample_Name")]
write.csv(cc_exprs_all, "cc_exprs_all.csv")
write.csv(cc_exprs_complete, "cc_exprs_complete.csv")
```

## do the same cell-types cluster together?

```{r}
result <- pls(cc_exprs_complete[, intersect(colnames(cc_exprs_complete), colnames(cc_mrna2))], cc_exprs_complete[, intersect(colnames(cc_exprs_complete), colnames(cc_cpg))])

plotIndiv(result)
plotVar(result, col=c("red", "black"), cutoff = 0.3, cex=rep(4, 2))
```

> common cell-type from both datasets cluster together


## Multiassay experiment (R)

## all samples

```{r}
mrna_exprs2 <- as.data.frame(t(mrna_exprs)) %>% 
  dplyr::mutate(sample = rownames(.))
cpg_exprs2 <- as.data.frame(t(cpg_exprs)) %>% 
  dplyr::mutate(sample = rownames(.))
mrna_cpg_exprs <- full_join(mrna_exprs2, cpg_exprs2, "sample")
rownames(mrna_cpg_exprs) <- mrna_cpg_exprs$sample

# mrna
mrna_exprs_all <- mrna_cpg_exprs[metadata_all$title, rownames(mrna_exprs)]
write.csv(mrna_exprs_all, "mrna_exprs_all.csv")

# cpg
cpg_exprs_all <- mrna_cpg_exprs[metadata_all$title, rownames(cpg_exprs)]
write.csv(cpg_exprs_all, "cpg_exprs_all.csv")

## check order of all sample
all(rownames(metadata_all) == rownames(mrna_exprs_all))
all(rownames(metadata_all) == rownames(cpg_exprs_all))
all(rownames(metadata_all) == rownames(cc_exprs_all))

## create MAE
mae_all <- MultiAssayExperiment(experiments = list(mrna = t(mrna_exprs_all), 
                                                   cpg = t(cpg_exprs_all), 
                                                   cc = t(cc_exprs_all)), 
                                colData = metadata_all)
mae_all@ExperimentList@metadata = list(mrna=mrna_ann, cpg=cpg_ann)
mae_all@metadata <- list("GSE" = "GSE71669")
saveHDF5MultiAssayExperiment(mae_all, dir = "gse71669_all", verbose = TRUE, replace = TRUE)
```

## complete cases

```{r}
# mrna
mrna_exprs_complete <- mrna_cpg_exprs[metadata_complete$title, rownames(mrna_exprs)]
write.csv(mrna_exprs_complete, "mrna_exprs_complete.csv")

# cpg
cpg_exprs_complete <- mrna_cpg_exprs[metadata_complete$title, rownames(cpg_exprs)]
write.csv(cpg_exprs_complete, "cpg_exprs_complete.csv")

## check order of all sample
all(rownames(metadata_complete) == rownames(mrna_exprs_complete))
all(rownames(metadata_complete) == rownames(cpg_exprs_complete))
all(rownames(metadata_complete) == rownames(cc_exprs_complete))

## create MAE
mae_complete <- MultiAssayExperiment(experiments = list(mrna = t(mrna_exprs_complete), 
                                                   cpg = t(cpg_exprs_complete), 
                                                   cc = t(cc_exprs_complete)), 
                                colData = metadata_complete)
mae_complete@ExperimentList@metadata = list(mrna=mrna_ann, cpg=cpg_ann)
mae_complete@metadata <- list("GSE" = "GSE71669")
saveHDF5MultiAssayExperiment(mae_complete, dir = "gse71669_complete", verbose = TRUE, replace = TRUE)

```

## save annotation files

```{r}
write.csv(mrna_ann, "mrna_ann.csv")
write.csv(cpg_ann, "cpg_ann.csv")
```

# Write to MuData

```{python}
import anndata
from anndata import AnnData
import mudata as md
from mudata import MuData
import pandas as pd

mrna_ann_pd = pd.read_csv("mrna_ann.csv", index_col=0)
mrna_ann_pd.index = mrna_ann_pd.index.map(str)
cpg_ann_pd = pd.read_csv("cpg_ann.csv", index_col=0)
cpg_ann_pd.index = cpg_ann_pd.index.map(str)

# all
metadata_all = pd.read_csv("metadata_all.csv", index_col=0)
metadata_all.index = metadata_all.index.map(str)
## mrna
mrna_exprs_pd_all = pd.read_csv("mrna_exprs_all.csv", index_col=0)
mrna_exprs_pd_all.index = mrna_exprs_pd_all.index.map(str)
mrna_anndata_all = AnnData(X = mrna_exprs_pd_all, obs = metadata_all, var = mrna_ann_pd, uns={"GSE": "GSE71669"})
## cpg
cpg_exprs_pd_all = pd.read_csv("cpg_exprs_all.csv", index_col=0)
cpg_exprs_pd_all.index = cpg_exprs_pd_all.index.map(str)
cpg_anndata_all = AnnData(X = cpg_exprs_pd_all, obs = metadata_all, var = cpg_ann_pd, uns={"GSE": "GSE71669"})
## cc
cc_exprs_pd_all = pd.read_csv("cc_exprs_all.csv", index_col=0)
cc_exprs_pd_all.index = cc_exprs_pd_all.index.map(str)
cc_anndata_all = AnnData(X = cc_exprs_pd_all, obs = metadata_all, uns={"GSE": "GSE71669"})
## merge data
mdata_all = MuData({"mrna": mrna_anndata_all, "cpg": cpg_anndata_all, "cc": cc_anndata_all})
mdata_all.write("../Python/gse71669_all.h5mu")

# complete
# all
metadata_complete = pd.read_csv("metadata_complete.csv", index_col=0)
metadata_complete.index = metadata_complete.index.map(str)
## mrna
mrna_exprs_pd_complete = pd.read_csv("mrna_exprs_complete.csv", index_col=0)
mrna_exprs_pd_complete.index = mrna_exprs_pd_complete.index.map(str)
mrna_anndata_complete = AnnData(X = mrna_exprs_pd_complete, obs = metadata_complete, var = mrna_ann_pd, uns={"GSE": "GSE71669"})
## cpg
cpg_exprs_pd_complete = pd.read_csv("cpg_exprs_complete.csv", index_col=0)
cpg_exprs_pd_complete.index = cpg_exprs_pd_complete.index.map(str)
cpg_anndata_complete = AnnData(X = cpg_exprs_pd_complete, obs = metadata_complete, var = cpg_ann_pd, uns={"GSE": "GSE71669"})
## cc
cc_exprs_pd_complete = pd.read_csv("cc_exprs_complete.csv", index_col=0)
cc_exprs_pd_complete.index = cc_exprs_pd_complete.index.map(str)
cc_anndata_complete = AnnData(X = cc_exprs_pd_complete, obs = metadata_complete, uns={"GSE": "GSE71669"})
## merge data
mdata_complete = MuData({"mrna": mrna_anndata_complete, "cpg": cpg_anndata_complete, "cc": cc_anndata_complete})
mdata_complete.write("gse71669_complete.h5mu")


```

# remove all csv files

```{bash}
rm *.csv
```

